Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-03-16 TEBenchmarkPaper/article.tex
Application

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application translates to Haskell and Isabelle, which also
  benefits from having definitions be strongly typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item Since it will act as our ground truth, the problem set should ideally
  contain every ``interesting'' conjecture involving its included
  definitions. Realistically, we should aim for each definition to appear in
  \emph{many} theorem statements; rather than each problem having unique
  definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

@@

@@

We propose that running times are compared using a Wilcoxon signed-rank
test~\cite{wilcoxon1945individual}, since it does not require that the data be
normally distributed. In our application we follow a variant of the Speedup-Test
protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the differences
are our use of the paired form of the Wilcoxon test, and the fact that each run
was performed on a different sample (to reduce sampling bias, at the expense of
accuracy).

Another benefit to such a paired difference test is its robustness to the
censoring caused by timeouts: upper-bounding the run time gives an upper-bound
to the observable difference, which biases our results \emph{towards} the null
hypothesis of indistinguishable performance.

To compare the recall we use the standard McNemar test for paired samples: each
ground-truth theorem is categorised as either being found only by the first
tool, only by the second, by both or by neither. Those found by both or neither
are ignored, and the tool-specific numbers are tested for marginal homogeneity
(i.e. whether their differences give rise to similar totals).

We cannot use McNemar's test to compare precision, since each tool may generate
different conjectures. Instead, we use Boschloo's form of Barnard's test to
compare the proportions of wanted and unwanted conjectures generated by each
tool.

@@

To determine our benchmarking parameters we ran some initial tests on both tools
for a few samples sized between 1 and 100, for an hour each. Most QuickSpec runs
either finished within 200 seconds or not at all, and sizes above 20 mostly
timed out. IsaCoSy mostly finished within 300 seconds on sample sizes up to 4,
but by size 8 was mostly timing out; its few successes above this took thousands
of seconds each, which we deemed infeasibly long.

>>>>

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item Since it will act as our ground truth, the problem set should
  \emph{ideally} contain every ``interesting'' conjecture involving its included
  definitions. More \emph{realistically}, we should aim for each definition to
  appear in \emph{many} theorem statements; rather than each problem having
  unique definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

@@

We prefer the latter, where each generated conjecture (wanted or not) is like a
coin toss whose bias is the precision; likewise, the ground truth theorems
(found or not) are like coin tosses biased by the recall. Note that this simple
model does not explain how many conjectures the tool decides to output for each
run.

@@

We propose that running times are compared using a Wilcoxon signed-rank
test~\cite{wilcoxon1945individual}, since it does not require that the data be
normally distributed. In $\S$\ref{sec:application} we follow a variant of the
Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
differences are our use of the paired form of the Wilcoxon test, and the fact
that each run was performed on a different sample (to reduce sampling bias, at
the expense of accuracy). This paired difference test is also robust to the
censoring caused by timeouts: upper-bounding the run time gives an upper-bound
to the observable difference, which biases our results in a conservative way
(\emph{towards} the null hypothesis of indistinguishable performance).

For simplicity, we suggest comparing quality by pooling together results from
all samples where both tools finished successfully. Recall can be compared using
the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
ground-truth theorem is categorised as either being found only by the first
tool, only by the second, by both or by neither. Those found by both or neither
are ignored, and the tool-specific numbers are tested for marginal homogeneity
(i.e. whether their differences give rise to similar totals).

We cannot use McNemar's test to compare precision, since each tool may generate
different conjectures. Instead, we can use Boschloo's form of Barnard's
test~\cite{lydersen2009recommended} to compare the proportions of wanted and
unwanted conjectures generated by each tool, assuming a binomial distribution
and conditioning on the number of generated conjectures (as if this were fixed
by design).

@@

To determine our benchmarking parameters we ran some initial tests on both tools
for a few samples sized between 1 and 100, for an hour each on our benchmarking
machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
above this took thousands of seconds each, which we deemed infeasibly long.
